Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete

### 2019-06-04 analysis/paper/paper2.Rmd
Final rewrite

#The appropriate packages are installed. Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots. The colour scheme in this paper is chosen from those that are suitable for colour-blind readers. The font Garmond is installed and stored.
library(ggpubr)
library(ineq)
library(reshape)
library(data.table)
library(extrafont)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)
library(plyr)
library(ggforce)
library(png)
font_import(paths="../paper/font",pattern="GARA")
loadfonts(device="win")

@@

For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA). Research into this once maligned period has grown apace in recent decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [see @Kotsonas2016; @Morris1997]. Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.

@@

As to the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses to elucidate social dynamics, settlement patterning, and systems of inter-regional communication across Crete and the wider Aegean. Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative research [@Bevan2013; @Knappett2011a; @Knappett2011]. The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research on Crete, where they have yet to make much of an impact [see @Kotsonas2011]. The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and studied, but a lack of systematic, quantitative comparisons between the cemeteries has left us with many intuited trends, some exceptional tombs, yet a general impression of un-patterned multiplicity.

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287]. Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism [@1950].

@@

## Heroes and Heirlooms: Object Biographies and Exceptional Burials

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986l; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence. Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA.

@@

Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data; only a few authors have pursued more quantative approaches. Cluster analyses have been employed by both Cavanagh [-@Coldstream1996: 653-57] and Antoniadis [-@Antoniadis2012: 193-97], the most sophisticated statistical techniques so far applied to the material. Antoniadis [-@Antoniadis2012: 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods. Yet even here the available data could be further interrogated (with, for instance, the number of imported items *per burial*). The present paper thus seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.

@@

The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a]. The first includes an entry for each individual ceramic vessel. Variables includes the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures. In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016]. In total, this database comprises just over 4540 individual artefacts.

@@

The final database concerns the tombs themselves. Each tomb was accorded its own entry, with variables including dating (both in relative and absolute terms), surviving dimensions, and the known and estimated number of interments, as well as a tally of the objects recorded in the other two databases. Finally, the total number of imports, imitations and antiques in each tomb was calculated.

@@

Ever more archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data, but the full methodologies underpinning our academic output. With the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [see @article]. In Marwick's words [-@Marwick2017: 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".

@@

```{r burialestimate, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}

@@

For most diachronic analyses, however, pithoi alone were taken as the best proxy for burial numbers. In most cases, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding what are presumably among the most heavily damaged and looted tombs.

@@

Another major concern was the temporal uncertainty implicit in the assigning of individual vessels to ceramic periods. A Protogeometric B vessel could, on our best estimates, date anywhere between 840 and 810 BC. This situation results in major challenges to the visualisation and analysis of such data. A simple, if coarse, solution is to use the midpoint of the range assigned to each vessel (i.e. 825 for our PGB example). This is used in a couple of instances in the present paper, for assigning vessels to broad, century-long windows.

```{r allpithoi,fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}

@@

A alternative approach, and the method employed extensively in this paper, is the use of so-called 'aoristic analysis' [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a]. Here, rather than give a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to one are then assigned for each vessel, per chunk of time, based on its accordant date range. So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of those three decades. These values for individual vessels can then simply be summed to produce tomb- or cemetery-level signatures. A particular attraction of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].

```{r aoristallcems, fig.dim = c(7, 4), dpi=300, fig.align='center', fig.cap="goes around here"}

@@

As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible. Though these methods present their own limitations (see @Crema2010, 1123-24), they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.

@@

##Introduction
The great advantage of the large, quantitative dataset employed here is that it permits investigation of several nested levels of potentially significant patterning. From individual burials, to collective tombs, to whole cemeteries, we have the opportunity to study and compare both the synchronic and diachronic variation across these various levels. In so doing, I first investigate the broad temporal trends in tomb construction, burial, and pottery deposition across both cemeteries, isolating a notable trend among a select group of tombs. I then examine these more closely, in an attempt to understand what sets them apart from the rest. Tomb dimensions, rates of burial, and changes in the composition of assemblages through time are all considered, to produce an account which speaks to the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.

@@

An obvious place to begin is with the chronology of the burials themselves. The rate and number of interments being made through time provides the background against which to consider all other developments.

```{r aoristbycem, dpi=300, fig.align="center", fig.cap="goesaround here.", fig.dim=c(8,4)}
KNC<-subset(tombs,tombs$cem_condensed=="KNC")

@@

Figure \@ref(fig:aoristbycem) presents the aoristic sums of pithoi (as a proxy for burials) and of all vessels, respectively. Despite differences in scale and timing, it is immediately apparent that Fortetsa and the KNC experieced similar chronological developments in burials and, to a slightly lesser degree, grave goods. In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th. Across all vessels, the developments are more volatile, but largely similar, especially regarding the late spike in burial activity. We might assume that these trends would be mirrored in the construction of new tombs, but this proves not to be the case.

```{r tombconstruction, fig.dim=c(7.5,4), dpi=300, fig.cap="goes around here"}

@@

```{r tombsinuse, fig.dim=c(7,4), dpi=300, fig.cap="goes around here"}

@@

```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=300, fig.cap="goes around here"}
tombs_11th<-subset(tombs,tombs$century_built=="11th")
tombs_10th<-subset(tombs,tombs$century_built=="10th")
tombs_9th<-subset(tombs,tombs$century_built=="9th")
tombs_8th<-subset(tombs,tombs$century_built=="8th")
tombs_7th<-subset(tombs,tombs$century_built=="7th")

@@

By dividing the tombs in use into the centuries during which they were built (Figure \@ref(fig:tombsinusebybuilddate)), we see a wave-like pattern with each cohort steadily going out of use as new ones were constructed. This would seem to accord with Cavanagh's (1996) suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use. We might expect, then, that it was the 8th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted.

```{r potsbydateandtombdate,  fig.dim=c(7,4), dpi=300, fig.cap="goes around here"}

@@

However, we see from Figure \@ref(fig:potsbydateandtombdate) that, although the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was tombs built back in the 9th and 10th which received the lion's share of the burials and associated ceramics. This is all the more notable when, returning to Figure \@ref(fig:tombsinusebybuilddate), we see that the number of 9th and 10th century tombs in use by this time was, in fact, declining.

@@

We are yet to draw out which older tombs precisely are responsible for the trend so far discussed. We might assume the tombs which became the focus of intense later burial activity would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.

```{r toptombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="goes around here"}

@@

The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of them did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)). In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries. All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials. Interestingly, Tombs X, G, Q, 219 and 283 ('Group B'), however, evince a different pattern, with their respective peaks coming around a century earlier, which, as seen in Figure \@ref(fig:aoristbycem), was another period of increased burial activity. So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via different routes. And, indeed, the differences between them may shed light on the late surge in burial activity here discussed.

@@

The indices employed so far – the distribution of artefacts between tombs, and the construction and use of the tombs through time – inform us of the temporal or aggregate features of the cemeteries, but tell us little of the tombs as they were built or first used. In attempting more synchronic comparisons, we shall first consider the physical dimensions of the tombs as built - a potentially rich context for the negotiation of social power, status and group membership.

```{r dimensionsboxplots, fig.dim=c(6,5),dpi=300, fig.align='center',fig.cap="goes around here"}

@@

In Figure \@ref(fig:dimensionsboxplots), we see that, through the centuries, tomb dimensions did vary, and sometimes consistently, as with the trend toward narrower *dromoi*. The 9th-century tombs, though, which include all but two of the 12 discussed above, do not particularly stand out. It does seem that chambers were, on average, larger in this period as well as exhibiting greater variety than at any other time. Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on construction would have been limited. The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this. It appears at any rate that they were gradually declining in size, albeit with some notable outliers.

```{r tombdimensions, fig.dim = c(7.5, 4), dpi=300, fig.align='center', fig.cap="goes around here"}

tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=cem_condensed))+geom_point()+
  theme_bw()+theme(legend.title=element_blank(),
        text=element_text(family="Garamond", size=12))+
            labs(x=bquote('Chamber Area ('*m^2*')'), y="Estimated Total Burials")+
               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=cem_condensed))+geom_point()+
  theme_bw()+theme(legend.title=element_blank(),
        text=element_text(family="Garamond", size=12))+
            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
               scale_color_brewer(palette = "Set2",labels=c(" KNC  "," Fortetsa"))+
                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

Furthermore, it seems that no clear relationship existed between the physical size of a tomb and its eventual buried population (see Figure \@ref(fig:tombdimensions)). True, over half of all tombs do not exceed 2.6 sq m in area, while nine of the ten tombs with the most burials do. But many of the very largest tombs are decidedly middling when it comes to burial numbers. Things are even less clear for the *dromos*; the largest two tombs by burial numbers also had exceptionally long *dromoi*, but the next eight tombs with the longest *dromoi* do not rank highly on interments.

It seems dubious, then, whether the scale of the tombs when built carried overt social meanings but, even if it did, this proved irrelevant to their later reuse. Several tombs had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns. Sometimes, urns from within the chamber were relocated to such niches to make space for new interments (e.g. T. TFT; Brock 1957: 3–4). The impression is of competition for inclusion into the sepulchral group, further implying the tombs’ original builders little anticipated their eventual popularity.

@@

Dimensions aside, we can still consider the early use of these tombs through comparisons between the scale and composition of their respective assemblages through time.

```{r tombscompthroughtime, fig.dim=c(7,4),dpi=300,fig.align='center',fig.cap="goes around here"}

@@

ggplot(pots9th10th,aes(tomb,fill=group))+geom_bar()+
  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12),
               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
                 guides(fill=guide_legend(title="Group"))+labs(x="Tomb", y="Vessel Count")+
                  scale_fill_brewer(palette = "Set2")+ 
                    scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
```

>>>>

#The appropriate packages are installed. Extrafont allows the use of non-standard fonts in R plots, ggplot2 is a powerful figure and graph plotting tool, and RColorBrewer includes a number of colour pallettes which can be applied to plots. The colour scheme in this paper is chosen from those that are suitable for colour-blind readers. The font Garmond is installed and stored.
library(ggpubr)
library(ineq)
library(reshape)
library(data.table)
library(extrafont)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)
library(plyr)
library(ggforce)
library(png)
library(ggrepel)
font_import(paths="../paper/font",pattern="GARA")
loadfonts(device="win")

@@

For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA). Research into this once maligned period has grown apace in recent decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [see @Kotsonas2016; @Morris1997]. Nonetheless, our knowledge of many EIA settlements remains underdeveloped in comparison with those of the second millennium BC.

@@

As to the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses to elucidate social dynamics, settlement patterning, and systems of inter-regional communication across Crete and the wider Aegean. Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative research [@Bevan2013; @Knappett2011a; @Knappett2011]. The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research on Crete, where they have yet to make much of an impact [see @Kotsonas2011]. The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and studied, but a lack of systematic, quantitative comparisons between them has left us with many intuited trends, some exceptional tombs, yet a general impression of un-patterned multiplicity.

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that its “remains [...] are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287]. Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@1950].

@@

## Heroes Abroad, Immigrants at Home

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence. Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA.

@@

Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data; only a few authors have pursued more quantative approaches. Cluster analyses have been employed by both Cavanagh [-@Coldstream1996: 653-57] and Antoniadis [-@Antoniadis2012: 193-97], the most sophisticated statistical techniques so far applied to the material. Antoniadis [-@Antoniadis2012: 172-76] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods. Yet even here the available data could be further interrogated (with, for instance, the number of imported items *per burial*). The present paper thus seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.

@@

The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a]. The first includes an entry for each individual ceramic vessel. Variables include the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures. In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016] respectively. In total, this database comprises just over 4540 individual artefacts.

@@

The final database concerns the tombs themselves. Each tomb was accorded its own entry, with variables including dating (both relative absolute), surviving dimensions, and the known and estimated number of interments, as well as tallies of the various artefact types recorded in the other two databases.

@@

Ever more archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data, but the full methodologies underpinning our academic output. With the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [@article]. In Marwick's words [-@Marwick2017: 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".

@@

```{r burialestimate, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}

@@

For most diachronic analyses, however, pithoi alone were taken as the best proxy for burial numbers. In most cases, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding the most heavily damaged and looted tombs.

@@

Another major concern was the temporal uncertainty implicit in the assigning of individual vessels to ceramic periods. A Protogeometric B vessel could, on our best estimates, date anywhere between 840 and 810 BC. This situation results in major challenges to the visualisation and analysis of such data. A simple, if coarse, solution is to take the midpoint of the range assigned to each vessel (i.e. 825 for our PGB example). This is used in a couple of instances in the present paper, for assigning vessels to broad, century-long windows.

```{r allpithoi,fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}

@@

A alternative approach, and the method employed extensively in this paper, is the use of so-called 'aoristic analysis' [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a]. Here, rather than give a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to one are then assigned for each vessel, per chunk of time, based on its accordant date range. So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of the three decades this encompasses. These values for individual vessels can then simply be summed to produce tomb- or cemetery-level signatures. A particular attraction of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].

```{r aoristallcems, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}

@@

As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible. Though these methods present their own limitations [@Crema2010, 1123-24], they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.

@@

##Introduction
The great advantage of the large, quantitative dataset employed here is that it permits investigation of several nested levels of potentially significant patterning. From individual burials, to collective tombs, to whole cemeteries, we have the opportunity to study and compare both the synchronic and diachronic variation across these various levels. In so doing, I first investigate the broad temporal trends in tomb construction, burial, and pottery deposition across both cemeteries, isolating a notable trend among a select group of tombs. I then examine these more closely, in an attempt to understand what sets them apart from the rest. Tomb dimensions, artefact distributions, and changes in the composition of assemblages through time are all considered, to produce an account which speaks to the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.

@@

An obvious place to begin is with the chronology of the burials themselves. The rate and quantity of interments being made through time provides the background against which to consider all other developments.

```{r aoristbycem, dpi=600, fig.align="center", fig.cap="goesaround here.", fig.dim=c(8,4)}
KNC<-subset(tombs,tombs$cem_condensed=="KNC")

@@

Figure \@ref(fig:aoristbycem) presents the aoristic sums of pithoi (as a proxy for burials) and of all vessels, respectively. Despite differences in scale and timing, it is immediately apparent that Fortetsa and the KNC underwent similar chronological developments with regard to burials and, to a slightly lesser degree, grave goods. In both cemeteries, the rate of cremation burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th. Across all vessels, the developments are more volatile, but largely similar, especially regarding the late spike in burial activity. We might assume that these trends would be mirrored in the construction of new tombs, but this proves not to be the case.

```{r tombconstruction, fig.dim=c(7.5,4), dpi=600, fig.cap="goes around here"}

@@

```{r tombsinuse, fig.dim=c(7,4), dpi=600, fig.cap="goes around here"}

@@

```{r tombsinusebybuilddate, fig.dim=c(7,4), dpi=600, fig.cap="goes around here"}
tombs_11th<-subset(tombs,tombs$century_built=="11th")
tombs_10th<-subset(tombs,tombs$century_built=="10th")
tombs_9th<-subset(tombs,tombs$century_built=="9th")
tombs_8th<-subset(tombs,tombs$century_built=="8th")
tombs_7th<-subset(tombs,tombs$century_built=="7th")

@@

By dividing the tombs in use into the centuries during which they were built (Figure \@ref(fig:tombsinusebybuilddate)), we see a wave-like pattern with each cohort steadily going out of use as new ones were constructed. This would seem to accord with Cavanagh's (1996: 659) suggestion that "[t]he length of time the tombs last seems [...] to reflect the duration of the social group which had access to the sepulchre", and that, if such access was lineally inehrited, frequent fission would result in limited trans-generational use. But note also that the total increase in active tombs during the 8th century belies the diminishing number of continuing 9th and 10th century tombs. We might expect, then, that it was the 8th century tombs coming into use that spurred the dramatic rise in burial rates previoulsy noted.

```{r potsbydateandtombdate,  fig.dim=c(7,4), dpi=600, fig.cap="goes around here"}

@@

However, we see from Figure \@ref(fig:potsbydateandtombdate) that, although the late 8th and early 7th centuries represented the apex of burial activity across both cemeteries, it was tombs built back in the 9th and 10th which received the lion's share of the burials and associated ceramics. This is all the more notable when, as seen in Figure \@ref(fig:tombsinusebybuilddate), they were, at this time, declining in number.

@@

We are yet to draw out which older tombs precisely are responsible for the trend so far discussed. We might assume the tombs which became the focus of such intense later burial activity would be those which yielded the most burials and grave goods when excavated and, happily, this supposition proves true.

```{r toptombs, fig.dim = c(7, 5), dpi=600, fig.align='center',fig.cap="goes around here"}

@@

The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of them did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)). In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries. All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.

```{r groups_a_and_b, fig.dim=c(7,5),dpi=600, fig.align='center',fig.cap="goes around here"}

tombs$group<-ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","A",
                   ifelse(tombs$tomb=="X"|tombs$tomb=="G"|tombs$tomb=="219"|tombs$tomb=="Q"|tombs$tomb=="283","B","C"))

GroupA<-subset(tombs,group=="A")

aorist_GroupA <- colSums(GroupA[158:207])
aorist_GroupA<-data.frame(aorist_GroupA)

colnames(aorist_GroupA)<-c("aoristic_sum")
row.names(aorist_GroupA)<-dates
aorist_GroupA<-data.frame(aorist_GroupA)
setDT(aorist_GroupA, keep.rownames = "decade")
aorist_GroupA$decade<-as.numeric(aorist_GroupA$decade)

aorist_GroupA$group<-"A"

GroupB<-subset(tombs,tombs$group=="B")

aorist_GroupB <-colSums(GroupB[158:207])
aorist_GroupB<-data.frame(aorist_GroupB)

colnames(aorist_GroupB)<-c("aoristic_sum")
row.names(aorist_GroupB)<-dates
aorist_GroupB<-data.frame(aorist_GroupB)
setDT(aorist_GroupB, keep.rownames = "decade")
aorist_GroupB$decade<-as.numeric(aorist_GroupB$decade)

aorist_GroupB$group<-"B"

aorist_groups<-rbind(aorist_GroupA,aorist_GroupB)

ggplot(aorist_groups,aes(decade,aoristic_sum,fill=group))+geom_col(position="dodge")+
  theme_bw()+theme_bw()+theme(text=element_text(family="Garamond", size=12))+
                labs(x="Date (Years BC)", y="Aoristic Sum")+
                  scale_fill_brewer(name="Group",palette = "Set2")+
                   scale_y_continuous(expand=c(0,0),limits=c(0,120))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1000,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

Interestingly, Tombs X, G, Q, 219 and 283 ('Group B'), however, evince a different pattern, with their respective peaks coming around a century earlier, which, as seen in Figure \@ref(fig:aoristbycem), was another period of increased burial activity. So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via quite different histories of use, a divergence made clear in Figure \@ref(fig:groups_a_and_b). It is notable that tombs assinged to each group are seen in both cemeteries, suggesting from the outset that the differences between these may speak to behaviours and social contexts operating across a significant portion of the Knossian community.

@@

Having addressed some broad diachronic trends, let us now attempt some more synchronic comparisons,   beginning with the physical dimensions of the tombs as built - a potentially rich context for the negotiation of social power, status and group membership. Perhaps the most bountiful tombs were also physically the largest, assertions of status and authority embodied in the man-power invested in their construction.

```{r dimensionsboxplots, fig.dim=c(6,5),dpi=600, fig.align='center',fig.cap="goes around here"}

@@

In Figure \@ref(fig:dimensionsboxplots), we see that, through the centuries, tomb dimensions did vary, and sometimes consistently, as with the trend toward narrower *dromoi*. The 9th-century tombs, though, which include all but two of the 12 discussed above, do not particularly stand out. It does seem that chambers were, on average, larger in this period, as well as exhibiting greater variety than at any other time. Then again, even the largest chambers are not exactly cavernous, and the extra labour expended on construction would have been limited. The *dromos* is a plausible area for conspicuous demonstrations of scale but, with their frequently poor preservation, we cannot be confident in assessing this. It appears at any rate that they were gradually declining in size, albeit with some notable outliers.

```{r tombgroupdimensions, fig.dim = c(7, 4), dpi=600, fig.align='center', fig.cap="goes around here"}

ggplot(subset(tombs,century_built=="9th" | century_built=="10th"),aes(dromos_length,chamber_area,color=group,label=tomb))+geom_point()+
  theme_bw()+theme(text=element_text(family="Garamond", size=12))+
  geom_label_repel(data=subset(tombs,group!="C"),aes(fill=group),color="white",show.legend = FALSE,family="Garamond")+
            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
               scale_color_brewer(palette = "Set2",name="Group")+
               scale_fill_brewer(palette = "Set2")+
                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,10))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))+coord_fixed()

Looking more specifically at the tombs of Groups A and B, against the backdrop of the other 10th and 9th century tombs, there are perhaps hints of differentiation. All of them fall within the top 50% based on chamber area, and three have exceptionally long *dromoi*. But, later destruction notwithstanding, there appears nothing systematic about the variation visible here. If the size of one's tomb was intended to evoke or assert ideas of wealth or status, this appears to have occurred on a largely independent basis.

```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="goes around here"}

tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=group))+geom_point()+
  theme_bw()+theme(text=element_text(family="Garamond", size=12))+
            labs(x="Chamber Area (sq m)", y="Estimated Total Burials")+
               scale_color_brewer(palette = "Set2",name="Group")+
                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=group))+geom_point()+
  theme_bw()+theme(text=element_text(family="Garamond", size=12))+
            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
               scale_color_brewer(palette = "Set2",name="Group")+
                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

What we can say is that no clear relationship existed between the physical size of a tomb and its eventual buried population (see Figure \@ref(fig:tombdimensions)). Many of the very largest tombs are decidedly middling when it comes to burial numbers, and vice versa. So whatever the scale of the tombs connoted when built, this proved irrelevant to their later reuse. Several had niches cut into their *dromoi*, or the *stomion* extended, to fit in later burial urns (including T. 292, 283 and 107; Cavanagh 1996: 658) and, sometimes, urns from within the chamber were relocated to these to make space for new interments (including T. 75 and TFT; ibid.: 658; Brock 1957: 3–4). The impression is of competition for inclusion into the sepulchral group, further confirming that the tombs’ original builders little anticipated their eventual popularity. Whatever led to Group A's curious later use was, it was not presaged by exceptional scale or physical grandeur.

@@

Dimensions aside, we can still investigate the contemporary use of these tombs through comparisons between the scale and composition of their respective assemblages through time.

```{r tombscompthroughtime, fig.dim=c(7,4),dpi=600,fig.align='center',fig.cap="goes around here"}

@@

ggplot(pots9th10th,aes(tomb,fill=group))+geom_bar()+
  theme_bw()+theme(legend.position="bottom",text=element_text(family="Garamond", size=12),
               axis.text.x = element_text(angle = 90,hjust=1,vjust=0.4))+
                 guides(fill=guide_legend(title="Group"))+labs(x="Tomb", y="Vessel Count")+
                  scale_fill_brewer(palette = "Set2")+ 
                    scale_y_continuous(expand=c(0,0))+facet_grid(~century,scales="free_x",space="free")
